Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-07 ms.tex
intro again

Knowledge of how $\dot{M}$ depends on the SMBH mass, $\Mbh$, and other
properties of the galactic nucleus informs key questions related to
the co-evolution of SMBHs and their host galaxies with cosmic time
(e.g.~\citealt{Kormendy&Ho13}; \citealt{Heckman&Best14}).  SMBH growth
in the low redshift universe is dominated by low mass black holes,
$M_{\bullet} \lesssim 10^{8}M_{\odot}$ (e.g.~\citealt{Heckman+04}), a
fact which is often attributed to the general trend of `cosmic
down-sizing' resulting from hierarchical structure growth (e.g,
\citealt{Gallo+08}).  However, the physical processes involved in the
growth of low mass blacks may be distinct from those operating at
higher SMBH masses and accretion rates in quasars.  A key question
here is whether SMBHs grow primarily by accreting gas fed in directly
from large scales (galactic or extra-galactic), or whether significant growth results also from
internal, secular processes such as local stellar mass loss in the nuclear region.

Obtaining a better understanding of the mechanisms regulating accretion onto quiescent SMBHs would have implications
for a variety of observats, including constraints on the occupation fraction of SMBHs in
low mass galaxies, and the potential existence of intermediate mass
black holes.  \citet{Miller+15} use the average correlation
between $\Mbh$ and the nuclear X-ray luminosities, $L_{X}$, of a sample
of elliptical galaxies to infer tentative evidence for a SMBH
occupation fraction less than unity in galaxies with stellar masses
$M_{\star} \lesssim 10^{10}M_{\odot}$ ($M_{\bullet} \lesssim
10^{7}M_{\odot}$).  This method relies on extrapolating a power-law
fit of the $L_X-\Mbh$ distribution to low values of $L_{\rm X}$ below
the detection threshold, a potentially worrisome assumption if different
physical processes control the accretion rates onto the lowest mass
SMBHs.

@@

The CNM density also influences the emission from stellar tidal
disruption events (TDEs), such as the high energy transient {\it
  Swift} J1644+57 (\citealt{Levan+11}; \citealt{Bloom+11},
\citealt{Burrows+11}; \citealt{Zauderer+11}).  This event was powered
by a transient relativistic jet, which produced synchrotron radio
emission as the jet material was decelerated by shock interaction with
the CNM of the previously quiescent SMBH
(\citealt{Giannios&Metzger11}; \citealt{Zauderer+11}).  Detailed
modeling of J1644+57 showed that the CNM density was significantly
lower than that measured surrounding Sgr A$^{\star}$
(\citealt{Metzger+12}; \citealt{Berger+12}).  However, a TDE jet
injected into a denser CNM would be decelerated more rapidly,
producing substantially different radio emission than in J1644+57.
Variations in the properties of the CNM could thus in principle help
explain why most TDEs appear to be radio quiet
(e.g.~\citealt{Bower+13}; \citealt{VanVelzen+13}).

@@

\citet{Ho:2009a} estimates the accretion rates onto a sample of
early-type galaxies both by using X-ray observations to determine the
Bondi accretion rate and using estimates for the mass loss rate of
evolved stars.  Both methods lead to the conclusion that the available
gas reservoir is more than sufficient to power the observed
low-luminosity AGN, assuming the standard $\sim$ 10 per cent radiative
efficiency for thin disk accretion.  Several independent strains of
evidence now suggest that low-luminosity AGN result from accretion
proceeding in a radiatively inefficient mode
(\citealt{Yuan&Narayan14}), due either to the advection of
gravitationally-released energy across the SMBH horizon
(e.g.~\citealt{Narayan&Yi95}) or due to a low efficiency for the gas
inflowing on large scales ultimately reaching the SMBH
(e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}).

@@

Another approach to determine the SMBH accretion rate, which we adopt
in this paper, is to directly calculate the density, velocity and
temperature profiles of the CNM using a physically motivated
hydrodynamic model.  Mass is injected to the nuclear environment via
stellar winds, while energy input results from several sources
including stellar winds, supernovae, and AGN feedback
(\citealt{Quataert:2004a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}).
Unlike previous works focusing on modeling individual galaxies, here
we systematically analyze the properties of the CNM across a sample of
galaxies with a range of SMBH masses, stellar density profiles, and
star formation histories (\citealt{WangMerritt:2004a}).

Past studies, employing multi-dimensional numerical hydrodynamics and
including variety of (parameterized) physical effects, have focused on
massive elliptical galaxies (e.g.~\citealt{Ciotti&Ostriker07};
\citealt{Ciotti+10}).  A key finding of these works is the periodic
development of cooling instabilities, which temporarily increase the
gas accretion rate until black hole feedback becomes sufficiently
important to shut off the flow and halt SMBH growth.  Here we instead
focus on time-independent models, neglecting radiative cooling.  This
approach allows us to systematically explore the relevant parameter
space and to derive simple analytic expressions that prove useful in
exploring whether cooling instabilties manifest across the expected
range of galaxy properties, or whether other (non-AGN) forms of
feedback can result in steady accretion.  Even if cooling instabilties
develop on galactic scales over longer timescales, e.g. $\sim$ Gyr, a
state of quasi-steady accretion may exist between these inflow events
on smaller radial scales comparable to the sphere of influence.  We
are thus motivated to explore whether steady flow is achieved within
the nucleus under any circumstances, or whether the gas content is
always building towards instability.
%A chief goal of our analysis is to provide a framework for interpreting the properties of quiescent galactic nuclei given the comparatively well understood physics of stellar winds and supernova heating.

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ can in many cases lie within a factor of few of the
nominal Bondi radius).  In the case of sufficiently weak heating,
however, no stagnation radius may exist at all, i.e. in principle the entire ISM of
the galaxy is feeding the SMBH.  Another goal of this work is to
assess under what conditions a stagnation radius exists, as this
question is intimately tied to that of cooling instability.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion.

>>>>

Knowledge of how $\dot{M}$ depends on the SMBH mass, $\Mbh$, and other
properties of the galactic nucleus informs key questions related to
the co-evolution of SMBHs and their host galaxies with cosmic time
(e.g.~\citealt{Kormendy&Ho13}; \citealt{Heckman&Best14}).  SMBH growth
in the low redshift universe is dominated by low mass black holes,
$M_{\bullet} \lesssim 10^{8}M_{\odot}$ (e.g.~\citealt{Heckman+04}), a
fact which is often attributed to the general trend of `cosmic
down-sizing' resulting from hierarchical structure growth (e.g,
\citealt{Gallo+08}).  However, the physical processes involved in growing low mass blacks may be distinct from those operating at
higher SMBH masses and accretion rates in quasars.  A key open question is whether SMBHs grow primarily by accreting gas fed in directly
from galactic or extra-galactic scales, or whether significant growth results also from
internal, secular processes such as local stellar mass loss in the nuclear region.

A better understanding of what mechanisms regulate accretion onto quiescent SMBHs would have implications
for a variety of observations, such as constraints on the occupation fraction of SMBHs in
low mass galaxies.  \citet{Miller+15} use the average relationship between the nuclear X-ray luminosities, $L_{X}$, of a sample of elliptical galaxies and the black hole mass to infer tentative evidence for a SMBH
occupation fraction less than unity in galaxies with stellar masses
$M_{\star} \lesssim 10^{10}M_{\odot}$ ($M_{\bullet} \lesssim
10^{7}M_{\odot}$).  This method relies on extrapolating a power-law
fit of the $L_X-\Mbh$ distribution to low values of $L_{\rm X}$ below
the detection threshold, a potentially problematic assumption if different
physical processes control the accretion rates onto the lowest mass
SMBHs.

@@

The gas density in galactic nuclei also influences the emission from stellar tidal
disruption events (TDEs), such as the high energy transient {\it
  Swift} J1644+57 (\citealt{Levan+11}; \citealt{Bloom+11},
\citealt{Burrows+11}; \citealt{Zauderer+11}).  This transient event was powered
by an impulsive relativistic jet, which produced synchrotron radio
emission as the jet material was decelerated by shock interaction with
the CNM of the previously quiescent SMBH
(\citealt{Giannios&Metzger11}; \citealt{Zauderer+11}).  Detailed
modeling of J1644+57 showed that the CNM density was much lower than that measured surrounding Sgr A$^{\star}$
(\citealt{Metzger+12}; \citealt{Berger+12}).  However, a TDE jet
injected into a denser CNM would be decelerated more rapidly,
producing substantially different radio emission than in J1644+57.
Variations in the properties of the CNM could help
explain why most TDEs appear to be radio quiet
(e.g.~\citealt{Bower+13}; \citealt{VanVelzen+13}).

@@

\citet{Ho:2009a} estimates the SMBH accretion rates in a sample of
early-type galaxies using X-ray observations to determine the
Bondi accretion rate and based on estimates for the mass loss rate of
evolved stars.  Both methods lead to the conclusion that the available
gas reservoir is more than sufficient to power the observed
low-luminosity AGN, assuming the standard $\sim$ 10 per cent radiative
efficiency for thin disk accretion.  Much
evidence now suggest that low-luminosity AGN result from accretion
proceeding in a radiatively inefficient mode
(\citealt{Yuan&Narayan14}), due either to the advection of
gravitationally-released energy across the SMBH horizon
(e.g.~\citealt{Narayan&Yi95}) or due to the low efficiency with which the inflowing gas ultimately reaching the viscinity of the SMBH
(e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}).

@@

Another approach to determine the SMBH accretion rate, which we adopt
in this paper, is to directly calculate the density, velocity and
temperature profiles of the CNM using a physically motivated
hydrodynamic model.  Mass is injected to the nuclear environment via
stellar winds, while energy input results from several sources
including stellar winds, supernovae, and AGN feedback
(\citealt{Quataert:2004a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}).
Unlike previous works, which focused primarily on modeling individual galaxies, here
we systematically analyze the CNM properties across a range of galaxy properties, including different SMBH masses, stellar density profiles, and
star formation histories (\citealt{WangMerritt:2004a}).

Past studies, employing multi-dimensional numerical hydrodynamics and
including variety of (parameterized) physical effects, have focused on
massive elliptical galaxies (e.g.~\citealt{Ciotti&Ostriker07};
\citealt{Ciotti+10}).  A key finding of these works is the periodic
development of cooling instabilities on galactic scales, which temporarily increase the
gas accretion rate towards the nucleus until feedback becomes sufficiently effective to shut off the flow and halt SMBH growth.  Here we instead
focus on time-independent models, neglecting radiative cooling.  This
approach allows us to systematically explore the relevant parameter
space and to derive simple analytic expressions that prove useful in
exploring whether cooling instabilties manifest across the expected
range of galaxy properties, or whether other (non-AGN) forms of
feedback can result in steady accretion.  Even if cooling instabilties
develop on galactic scales over longer timescales, e.g. $\sim$ Gyr, a
state of quasi-steady accretion may exist between these inflow events
on smaller radial scales comparable to the sphere of influence.  We
are thus motivated to explore whether steady flow is achieved within
the nucleus under any circumstances, or whether the gas content is
always building towards instability.
%A chief goal of our analysis is to provide a framework for interpreting the properties of quiescent galactic nuclei given the comparatively well understood physics of stellar winds and supernova heating.

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ can in many cases lie within a factor of few of the
nominal Bondi radius).  When heating is sufficiently weak, however, no stagnation radius may exist at all, i.e. the entire ISM of the galaxy is flowing towards the SMBH.  A key goal of this work is to
assess under what conditions a stagnation radius exists, as this
question is intimately tied to that of cooling instability.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion.
